### 2018-03-09 submission/manuscript.tex
discussion edits with pat comments in mind

Although we expected that the full OTU models would perform the best at
classifying individuals with and without carcinomas, our observations
suggest that both the full and significant OR taxa models performed
equally well. These results suggest that lower level classification to
species and strain may not add extra useful information with respect to
prediction models. This has been suggested in previous literature where
metagenomics did not perform better than 16S rRNA gene sequencing data
at classifying individuals with normal colons and those with carcinomas
(30). One possible reason as to why lower level classification may not
result in better models is that the communities are patchy and higher
level taxonomic information pools some of this patchiness, allowing for
better prediction models. There may also be a fair bit of data
redundancy within models that utilize more of the community. An example
of this redundancy would be when we trained models on one study and
tested it on the other studies and the AUCs of the models created with
the select OR taxa performed as well as full taxa models {[}Figure
6B{]}.

Our observations also suggest that a small collection of taxa can
classify disease as well as full OTU-based models but that these taxa
individually perform quite poorly {[}Figure 3{]}. This result supports
the contention that there might be redundancy of function amongst the
taxa included in the significant OR models. As an example, multiple
different microbes could be similarily stimulating the activation of
inflammatory pathways and by doing so exacerbate disease progression.
Multiple reports within the literature have found that different
bacteria, such as \emph{Escherichia coli} and \emph{Fusobacterium
nucleatum}, can similarily worsen inflammation in mouse models of
tumorigenesis (5, 6, 21). Although the inflammatory taxa were patchy in
their importance and presence across studies those that were not
typically associated with inflammation were consistently important for
every study {[}Figure 6{]}. The loss of these taxa (\emph{Ruminococcus}
and \emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia}
in unmatched tissue) is particularly interesting because many are
commonly thought to be beneficial due to their involvement in production
of short chain fatty acids (31--33).

The adenoma models as a whole performed poorly in classifying
individuals with and without adenomas. This outcome is not inconsistent
with what has been published previously (27, 34). However, the modeling
results are at odds with results obtained in Baxter, \emph{et al.} (12).
There are some major differences between the models generated in this
meta-analysis and what was used in Baxter, \emph{et al}. First, the
prevoius report's models investigated the classification of lesions
(individuals with adenoma or carcinoma) and not adenoma alone. The
Baxter, \emph{et al.} models also contained Fecal Immunoglobulin Test
data while our meta-analysis models only contained 16S rRNA gene
sequencing data. Although being able to classify individuals with
adenomas is important, the classification of advanced adenomas is a more
clinically meaningful diagnostic tool (i.e.~those that are at highest
risk of progressing to a carcinoma). It is possible that we might have
been able to detect differences in the bacterial community if advanced
adenomas were separated from adenomas but that data was not available
for the majority of studies analyzed. Additionally, the initial changes
to the bacterial community could be focal to where the initial adenoma
develops and would not be easily assessed with a fecal sample.

Although stool represents an easy and less invasive way to assess risk,
it is not clear how well this sample reflects adenoma- and carcinoma-
tissue associated microbial communities. The colon tissue-based studies
did not provide a clearer understanding of how the microbiota may be
associated with tumors. Generally, the full OTU-based models of
unmatched and matched colon tissue samples were concordant with stool
samples showing that GI resident microbes were the most prevalent in the
top 10 most important variables across study {[}Figure S3{]}.
\emph{Fusobacterium} was not consistently identified across studies and
this could be due to both a small number of studies and a small sample
size within these studies. The majority of the colon tissue-based
results were also study specific with many of the top 10 taxa being
present only in a single study. Additionally, the presence of genera
associated with contamination, within the top 10 most important
variables for the genera and OTU models is worrying (e.g.
\emph{Novosphingobium}, \emph{Acidobacteria Gp2}, \emph{Sphingomonas},
etc. (35)). The low bacterial biomass of tissue samples coupled with
potential contamination and small sample sizes could explain why these
results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC (22, 24). There are reports that
\emph{Bacteroides fragilis}, positive for the enterotoxigenic gene, are
found at specific locations along the colon but the samples we were able
to use in this meta-analysis could not identify these types of
differences (36). Additionally, since we are limited in our aggregation
of the data to the genus level, it is not possible to clearly delineate
which species are contributing to overall disease progression. Our
observations are not inconsistent with the previous literature on either
\emph{Bacteroides fragilis} or \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus}. As an example, the stool-based full community
models consistently identified the \emph{Bacteroides} taxa to be
important model components across studies. This suggests that even
though \emph{Bacteroides} may not increase the OR of individuals having
an adenoma or carcinoma and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC.

Despite these limitations to the findings, meta-analyses are a useful
tool in microbiota research because they can both validate existing
research and make new discoveries by pooling many independent
investigations together. Yet, it is still difficult to perform these
studies because of inaccessible 16S sequencing data, missing or vague
metadata (e.g.~which samples are carcioma and which are not), varying
sequence quality, and multiple small data sets. Better attention to
these specific problems could help to increase the reproducibility and
replicability of microbiota studies and make it easier to perform these
crucial meta-analyses. Moving forward, meta-analyses will be important
tools to help aggregate and find commonalities across studies when
investigating the microbiota in the context of a specific disease and
more are needed (28, 37--39).

By aggregating together a large collection of studies analyzing both
fecal and colon tissue samples, we are able to provide evidence
supporting the importance of the bacterial community in carcinoma
tumors. Although further validation of the biomarkers presented here
need to be undertaken, the replicability of the AUC of a specific
collection of taxa across multiple studies suggests a strong potential
for the use of the microbiota as a risk stratification tool for
individuals with carcinomas.

@@

\textbf{\emph{Datasets:}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (40) and
Vogtmann, et al. (41). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
excluded because they did not have publicly available sequences or did
not have metadata in which the authors were able to share. We were able
to obtain sequence data and metadata from the following studies: Ahn,
\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (42), Lu, \emph{et
al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
(43) study was excluded because their 90 individuals were contained
within the larger Baxter study (12). The Kostic study was excluded
because after we processed the sequences, all of the case samples had
100 or fewer sequences. The final analysis included 14 studies (Tables 1
and 2). There were seven studies with only fecal samples (Ahn, Baxter,
Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
fecal and tissue samples (Chen and Flemer). After curating the
sequences, 1737 stool samples and 492 tissue samples remained in the
analysis {[}Tables 1 and 2{]}.

\textbf{\emph{Sequence Processing:}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n = 4,
(17, 23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) (44) using the default quality filtering methods for both 454
and Illumina sequence data. If it was not possible to use the defaults
because the sequences were trimmed too much, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (45). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (46)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (47)).

\textbf{\emph{Community analysis:}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (48) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (49). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis:}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. Within the training
dataset, 10-fold cross validation (5-fold cross validation for small
datasets) was used to build a model that was then evaluated on the
testing set. For the models constructed using the taxa with significant
ORs, the default mtry setting was used to train the model and this model
was tested on the other datasets in the meta-analysis. The reported AUC
values are the AUCs for the application of the model on the test sets.
For the OTU-based models, the dataset was split into training (80\% of
samples) and testing (20\%) sets and 10-fold cross validation (5-fold
cross validation for small datasets) on the training set was used to
generate the model for the testing set. The original 80/20 split and
fitting was repeated 100 times and the average AUC from these 100
repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
the importance of each taxon to the overall model was used to rank the
taxa used in each model. For all models, the default setting used was
\(\sqrt{p}\), where p is all the variables used in the respective model.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (50). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (51) and then
Z-score normalized using the car (v2.1.6) package (52). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models created using the lme4 (v1.1.15) package (53) to
correct for study, repeat sampling of individuals (tissue only), and 16S
hyper-variable region used. Odds ratios (OR) were analyzed using both
the epiR (v0.9.93) and metafor (v2.0.0) packages (54, 55) by assessing
how many individuals with and without disease were above and below the
overall median value within each specific study. OR significance testing
utilized the chi-squared test. Diversity differences measured by the
Bray-Curtis index utilized the creation of distance matrix and testing
with PERMANOVA executed with the vegan (v2.4.5) package (56). Random
Forest models were built using both the caret (v6.0.78) and randomForest
(v4.6.12) packages (57, 58). All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages (59, 60).

@@

\hypertarget{ref-Hannigan2017}{}
30. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
Ruffin}, \textbf{C. C. Koumpouras}, and \textbf{P. D. Schloss}. 2017.
Diagnostic potential \& the interactive dynamics of the colorectal
cancer virome. Cold Spring Harbor Laboratory.

\hypertarget{ref-Venkataraman2016}{}
31. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
Schmidt}, \textbf{C. Waldron}, \textbf{K. R. Theis}, and \textbf{T. M.
Schmidt}. 2016. Variable responses of human microbiomes to dietary
supplementation with resistant starch. Microbiome. Springer Nature
\textbf{4}.

\hypertarget{ref-Herrmann2018}{}
32. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
Reichert-Grimm}, \textbf{S. Weis}, \textbf{C. Riedel}, \textbf{D.
Rosendale}, \textbf{H. Stoklosinski}, \textbf{M. Hunt}, and \textbf{M.
Egert}. 2018. In vivo assessment of resistant starch degradation by the
caecal microbiota of mice using RNA-based stable isotope probingA
proof-of-principle study. Nutrients. MDPI AG \textbf{10}:179.

\hypertarget{ref-Reichardt2017}{}
33. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
\textbf{F. M. Farquharson}, \textbf{D. Wefers}, \textbf{M. Bunzel},
\textbf{S. H. Duncan}, \textbf{J. E. Drew}, \textbf{L. M. Williams},
\textbf{G. Milligan}, \textbf{T. Preston}, \textbf{D. Morrison},
\textbf{H. J. Flint}, and \textbf{P. Louis}. 2017. Specific
substrate-driven changes in human faecal microbiota composition contrast
with functional redundancy in short-chain fatty acid production. The
ISME Journal. Springer Nature \textbf{12}:610--622.

\hypertarget{ref-Sze2017}{}
34. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
Normalization of the microbiota in patients after treatment for colonic
lesions. Microbiome. Springer Nature \textbf{5}.

\hypertarget{ref-Salter_contamination_2014}{}
35. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
critically impact sequence-based microbiome analyses. BMC Biology.
Springer Nature \textbf{12}.

\hypertarget{ref-Purcell2017}{}
36. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
2017. Colonization with enterotoxigenic bacteroides fragilis is
associated with early-stage colorectal neoplasia. PLOS ONE. Public
Library of Science (PLoS) \textbf{12}:e0171602.

\hypertarget{ref-Sze2016}{}
37. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
signal in the noise: Revisiting obesity and the microbiome. mBio.
American Society for Microbiology \textbf{7}:e01018--16.

\hypertarget{ref-Walters2014}{}
38. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
2014. Meta-analyses of human gut microbes associated with obesity and
IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.

\hypertarget{ref-Finucane2014}{}
39. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
Public Library of Science (PLoS) \textbf{9}:e84689.

\hypertarget{ref-keku_gastrointestinal_2015}{}
40. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
\textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology \textbf{308}:G351--G363.

>>>>

We performed a meta-analysis to identify and validate microbiome-base
biomarkers that could be used to classify individuals as having normal
colons or colonic tumors using fecal or tissue samples. To our surprise,
Random Forest classification models constructed to differentiate
individuals with normal colons from those with carcinomas using a subset
of the community performed well relative to models constructed using the
full communities. When we applied the models trained on each dataset to
the other datasets in our study, we found that the models trained using
the subset of the communities performed better than those using the full
communities. These models were trained using data in which sequences
were assigned to bacterial taxa using a classifier that typically
assigned sequences to the family or genus level. When we attempted to
improve the specificity of the classification by using an OTU-based
approach the resulting models performed as well as those constructed
using coarse taxonomic assignments. These results are significant
because they strengthen the growing literature indicating a role of the
microbiome in tumorigenesis (9) and as a potential tool as a
non-invasive diagnostic and for assessing risk of disease and recurrence
(12, 30).

These results suggested that fine scale classification of sequences into
OTUs does not improve our classification models. This has been suggested
in previous literature where shotgun metagenomic data did not perform
better than 16S rRNA gene sequencing data in classifying individuals
with normal colons and those with carcinomas (31). We hypothesize that
fine scale classification may not result in better classification
because distribution of microbiota between individuals is patchy. In
contrast, models using coarser taxonomic assignments will pool the fine
scale diversity and resulting in less patchiness and better
classification. Furthermore, the ability of models trained using a
subset of the community to outperform those using the full community
when testing the models on the other datasets may also be a product of
the patchiness of human-associated microbiota. The models based on the 8
taxa that had significant ORs used taxa that were found in every study
and tended to have higher relative abundances. Similar to the OTU-based
models, those models based on the full community taxonomy assignments
were still sensitive to the patchy distribution of taxa. Regardless, it
is encouraging that a collection of 8 taxa could reliably classify
individuals as having carcinomas considering the differences in cohorts,
DNA extraction procedures, regions of the 16S rRNA gene, and sequencing
methods.

When used separately to classify individuals with carcinomas, the taxa
with significant ORs could not reliably classify individuals {[}Figure
3{]}. This result further supports the hypothesis that
carcinoma-associated microbiota have a patchy distribution. Two
individuals may have had the same classification, based on the relative
abundance of different populations within this group of 8 taxa. Although
these results only reflect associations with disease, it is tempting to
hypothesize that the patchiness represents distinct mechanisms of
exacerbating tumorigenesis or that multiple taxa have the same mechanism
of exacerbating tumorigenesis. For example, strains of \emph{Escherichia
coli} and \emph{Fusobacterium nucleatum} have been shown to worsen
inflammation in mouse models of tumorigenesis (5, 6, 21). In contrast to
the patchiness of the taxa that were positively associated with
carcinomas, potentially beneficial taxa had a more consistent
association {[}Figure 6{]}. This result was particularly interesting
because members of these taxa (i.e. \emph{Ruminococcus} and
\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
tissue) are thought to be beneficial due to their involvement in
production of anti-inflammatory short chain fatty acids (32--34).

All of the adenoma classification models performed poorly. This result
is not inconsistent with previous studies (27, 30). However, the
classification results are at odds with results of the multitarget
microbiota test (MMT) from Baxter, \emph{et al.} (12) who observed an
AUC of 0.755 when applied to individuals with adenomas. There are two
major differences between the models generated in this meta-analysis and
that analysis. The MMT attempted to classify individuals as having a
normal colon or having colonic lesions (i.e.~adenomas or carcinomas) and
not adenomas alone. Further, the MMT incorporated fecal immunoglobulin
test (FIT) data while our models only used 16S rRNA gene sequencing
data. Because FIT data were not available for the other studies in our
meta-analysis, it was not possible to validate the MMT approach. The
ability to differentiate between individuals with and without adenomas
is an important problem since early detection of tumors is critical.
However, it is possible that we might have been able to detect
differences in the bacterial community if individuals with non-advanced
and advanced adenomas were separated. This is a clinically relevant
distinction since advanced adenomas are at highest risk of progressing
to a carcinoma. The initial changes of the microbiota during
tumorigenesis could be focal to where the initial adenoma develops and
would not be easily assessed using fecal samples from an individual with
non-advanced adenomas. Unfortunately, distinguishing between individuals
with advanced and non-advanced adenomas was not possible in our
meta-analysis since the studies did not provide the clinical data needed
to make that distinction.

Stool samples represent a non-invasive approach to assessing the
structure of the gut microbiota and are potentially useful for
diagnosing individuals as having colonic tumors. However, they do not
reflect the structure of the mucosal microbiota (35). Regardless, the
taxa that were the most important in the stool-based models overlapped
with those from the models trained using the data from unmatched and
matched colon tissue samples {[}Figure S3{]}. Mucosal biopsies are
preferred for focused mechanistic studies and have offered researchers
the opportunity to sample healthy and diseased tissue from the same
individuals (i.e.~matched) using each individual as their own control or
in a cross sectional design (i.e.~unmatched). Because obtaining these
samples is invasive, carries risks to the individual, and is expensive,
studies investigating the structure of the mucosal microbiota generally
have a limited number of participants. Thus, it was not surprising that
tissue-based studies did not provide clearer associations between the
mucosal microbiota and the presence of tumors. Interestingly,
Fusobacterium, which has received increased recent attention for its
potential role in tumorigenesis (6) was not consistently identified
across the studies in our meta-analysis. This could be due to the
relatively small number of individuals in the limited number of studies.
The classification models trained using the tissue-based data performed
well when tested with the training data (Figure S4), but performed
poorly when tested on the other tissue-associated datasets (Figure S5).
Disturbingly, taxa that are commonly associated with reagent
contamination (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
\emph{Sphingomonas}, etc.) were detected within the tissue datasets.
Such contamination is common in studies where there is relatively low
bacterial biomass (36). The lack of replication among the tissue-based
biomarkers may be a product of the relatively small number of studies
and individuals per study and possible reagent contamination.

Among our stool samples, we failed to identify several notable
populations that are commonly associated with carcinomas including an
enterotoxigenic strain of \emph{Bacteroides fragilis} (ETBF) and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} (22, 24).
ETBF have been found in tumors in the proximal colon where they tend to
form biofilms (20, 37). Considering DNA from bacteria that are more
prevalent in the proximal colon may be degraded by the time it leaves
the body, it is not surprising that we failed to identify a significant
OR for Bacteroides with carcinomas. In addition, since our approach
could only classify sequences to the genus level and there are likely
multiple Bacteroides populations in the colon, it is possible that
sequences from ETBF and non-oncogenic Bacteroides were pooled. This
would then reduce the OR between Bacterioides and whether an individual
had carcinomas. It is also necessary to distinguish between populations
that are biomarkers for a disease and those that are known to cause
disease. The former may also have a causative role in the disease.
Although the latter have been shown to have a causative role, they may
appear at low relative abundance, be found in specific locations, or may
have a highly patchy distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can
demonstrate whether a result can be replicated and facilitate new
discoveries by pooling multiple independent investigations. There have
been several meta-analyses similar to this study that have sought
biomarkers for obesity (38--40), inflammatory bowel disease (39), and
colorectal cancer (28). Considering microbiome research is particularly
prone to hype and overgeneralization of results, these analyses are
critical. For example, previous meta-analyses have demonstrated that
there are no clear fecal biomarkers for obesity (39, 40). Such
meta-analyses are difficult to perform because the underlying 16S rRNA
gene sequence data are not publicly available, metadata are missing,
incomplete, or vague, sequence data are of poor quality or derived by
non-standard approaches, and the original studies were significantly
underpowered. Reluctance to publish negative results (i.e.~the ``file
drawer effect'') is also likely to skew our understanding of the
relationship between microbiota and disease. Better attention to these
specifics will increase the reproducibility and replicability of
microbiota studies and make it easier to perform these crucial
meta-analyses. Moving forward, meta-analyses will be important tools to
help aggregate and find commonalities across studies when investigating
the microbiota in the context of a specific disease (28, 38--40).

Our meta-analysis suggests a strong association between the gut
microbiota and colon tumorigenesis. By aggregating the results from
studies that sequenced the 16S rRNA gene from fecal and tissue samples,
we are able to provide evidence supporting the use of microbial
biomarkers to diagnose the presence of colonic tumors. Further
development of microbial biomarkers should focus on including other
biomarkers (e.g.~FIT), better categorizing of people with adenomas, and
expanding datasets to include larger numbers of individuals. Based on
prior research into the physiology of the biomarkers we identified, it
is likely that they have a causative role in tumorigenesis. Their patchy
distribution across individuals suggests that there are either multiple
mechanisms causing disease or a single mechanism (e.g.~inflammation)
that can be mediated by multiple, diverse bacteria.

@@

\textbf{\emph{Datasets:}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (41) and
Vogtmann, et al. (42). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
excluded because they did not have publicly available sequences or did
not have metadata in which the authors were able to share. We were able
to obtain sequence data and metadata from the following studies: Ahn,
\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
(44) study was excluded because their 90 individuals were contained
within the larger Baxter study (12). The Kostic study was excluded
because after we processed the sequences, all of the case samples had
100 or fewer sequences. The final analysis included 14 studies (Tables 1
and 2). There were seven studies with only fecal samples (Ahn, Baxter,
Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
fecal and tissue samples (Chen and Flemer). After curating the
sequences, 1737 stool samples and 492 tissue samples remained in the
analysis {[}Tables 1 and 2{]}.

\textbf{\emph{Sequence Processing:}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n = 4,
(17, 23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) (45) using the default quality filtering methods for both 454
and Illumina sequence data. If it was not possible to use the defaults
because the sequences were trimmed too much, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (46). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (47)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (48)).

\textbf{\emph{Community analysis:}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (49) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (50). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis:}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. Within the training
dataset, 10-fold cross validation (5-fold cross validation for small
datasets) was used to build a model that was then evaluated on the
testing set. For the models constructed using the taxa with significant
ORs, the default mtry setting was used to train the model and this model
was tested on the other datasets in the meta-analysis. The reported AUC
values are the AUCs for the application of the model on the test sets.
For the OTU-based models, the dataset was split into training (80\% of
samples) and testing (20\%) sets and 10-fold cross validation (5-fold
cross validation for small datasets) on the training set was used to
generate the model for the testing set. The original 80/20 split and
fitting was repeated 100 times and the average AUC from these 100
repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
the importance of each taxon to the overall model was used to rank the
taxa used in each model. For all models, the default setting used was
\(\sqrt{p}\), where p is all the variables used in the respective model.
Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
mtry (51).

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (52). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (53) and then
Z-score normalized using the car (v2.1.6) package (54). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models created using the lme4 (v1.1.15) package (55) to
correct for study, repeat sampling of individuals (tissue only), and 16S
hyper-variable region used. Odds ratios (OR) were analyzed using both
the epiR (v0.9.93) and metafor (v2.0.0) packages (56, 57) by assessing
how many individuals with and without disease were above and below the
overall median value within each specific study. OR significance testing
utilized the chi-squared test. Diversity differences measured by the
Bray-Curtis index utilized the creation of distance matrix and testing
with PERMANOVA executed with the vegan (v2.4.5) package (58). Random
Forest models were built using both the caret (v6.0.78) and randomForest
(v4.6.12) packages (59, 60). All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages (61, 62).

@@

\hypertarget{ref-Sze2017}{}
30. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
Normalization of the microbiota in patients after treatment for colonic
lesions. Microbiome. Springer Nature \textbf{5}.

\hypertarget{ref-Hannigan2017}{}
31. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
Ruffin}, \textbf{C. C. Koumpouras}, and \textbf{P. D. Schloss}. 2017.
Diagnostic potential \& the interactive dynamics of the colorectal
cancer virome. Cold Spring Harbor Laboratory.

\hypertarget{ref-Venkataraman2016}{}
32. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
Schmidt}, \textbf{C. Waldron}, \textbf{K. R. Theis}, and \textbf{T. M.
Schmidt}. 2016. Variable responses of human microbiomes to dietary
supplementation with resistant starch. Microbiome. Springer Nature
\textbf{4}.

\hypertarget{ref-Herrmann2018}{}
33. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
Reichert-Grimm}, \textbf{S. Weis}, \textbf{C. Riedel}, \textbf{D.
Rosendale}, \textbf{H. Stoklosinski}, \textbf{M. Hunt}, and \textbf{M.
Egert}. 2018. In vivo assessment of resistant starch degradation by the
caecal microbiota of mice using RNA-based stable isotope probingA
proof-of-principle study. Nutrients. MDPI AG \textbf{10}:179.

\hypertarget{ref-Reichardt2017}{}
34. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
\textbf{F. M. Farquharson}, \textbf{D. Wefers}, \textbf{M. Bunzel},
\textbf{S. H. Duncan}, \textbf{J. E. Drew}, \textbf{L. M. Williams},
\textbf{G. Milligan}, \textbf{T. Preston}, \textbf{D. Morrison},
\textbf{H. J. Flint}, and \textbf{P. Louis}. 2017. Specific
substrate-driven changes in human faecal microbiota composition contrast
with functional redundancy in short-chain fatty acid production. The
ISME Journal. Springer Nature \textbf{12}:610--622.

\hypertarget{ref-Flynn_preprint_2017}{}
35. \textbf{Flynn, K. J.}, \textbf{M. T. Ruffin}, \textbf{D. K.
Turgeon}, and \textbf{P. D. Schloss}. 2017. Spatial variation of the
native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory.

\hypertarget{ref-Salter_contamination_2014}{}
36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
critically impact sequence-based microbiome analyses. BMC Biology.
Springer Nature \textbf{12}.

\hypertarget{ref-Purcell2017}{}
37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
2017. Colonization with enterotoxigenic bacteroides fragilis is
associated with early-stage colorectal neoplasia. PLOS ONE. Public
Library of Science (PLoS) \textbf{12}:e0171602.

\hypertarget{ref-Sze2016}{}
38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
signal in the noise: Revisiting obesity and the microbiome. mBio.
American Society for Microbiology \textbf{7}:e01018--16.

\hypertarget{ref-Walters2014}{}
39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
2014. Meta-analyses of human gut microbes associated with obesity and
IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.

\hypertarget{ref-Finucane2014}{}
40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
Public Library of Science (PLoS) \textbf{9}:e84689.
